---
title: "Flight Delays Analysis using R"
author: "Manish Patwardhan"
date: "2026-01-27"
output: word_document
---

# Step 1: Install Required Packages

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("readxl")
library(readxl)
```

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("ggplot2")
library(ggplot2)
```

```{r}
options(repos = c(CRAN = "https://cloud.r-project.org"))
install.packages("dplyr")
library(dplyr)
```

# Step 2: Read the Dataset + Understand Data

```{r}
# Load dataset
flight_data <- read_excel("flightdelays.xlsx")
```

```{r}
# Viwe first rows
head(flight_data)
```

```{r}
# Structure of dataset
str(flight_data)
```

```{r}
# Dimensions (rows, columns)
dim(flight_data)
```

```{r}
# Check missing values
colSums(is.na(flight_data))
```

```{r}
# Dimensions (rows, columns)
dim(flight_data)
```

# Step 3: Summary of Descriptive Statistics

```{r}
# Summary of Descriptive Statistics
summary(flight_data)
```

```{r}
# Histograms (Relationships)
ggplot(flight_data, aes(x = schedtime)) +
  geom_histogram() +
  ggtitle("Histogram of Scheduled Time")
```

# Step 4: Histograms (Relationships)

```{r}
# Scheduled Time
ggplot(flight_data, aes(x = schedtime)) +
  geom_histogram() +
  ggtitle("Histogram of Scheduled Time")
```

```{r}
# Carrier
ggplot(flight_data, aes(x = carrier)) +
  geom_bar() +
  ggtitle("Flights by Carrier")
```

```{r}
# Destination
ggplot(flight_data, aes(x = dest)) +
  geom_bar() +
  ggtitle("Flights by Destination")
```

```{r}
# Origin
ggplot(flight_data, aes(x = origin)) +
  geom_bar() +
  ggtitle("Flights by Origin Airport")
```

```{r}
# Weather (0=On time, 1=Delayed)
ggplot(flight_data, aes(x = weather)) +
  geom_bar() +
  ggtitle("Weather Impact on Delays")
```

```{r}
# Day of Week
ggplot(flight_data, aes(x = dayweek)) +
  geom_bar() +
  ggtitle("Flights by Day of Week")
```

# Step 5: Scatter Plot (On Time vs Delayed Flights)

```{r}
ggplot(flight_data, aes(x = distance, y = deptime, color = as.factor(delay))) +
  geom_point() +
  ggtitle("Scatter Plot: Distance vs Departure Time") +
  labs(color = "Delay Status")
```

# Step 6: Box Plot (Delay by Day of Month)

```{r}
ggplot(flight_data, aes(x = as.factor(daymonth), y = deptime)) +
  geom_boxplot() +
  ggtitle("Departure Time Distribution Across Month Days")
```

# Step 7: Define Hours of Departure

```{r}
flight_data$dep_hour <- floor(flight_data$deptime / 100)

head(flight_data$dep_hour)
```

# Step 8: Categorical Representation Using Table

```{r}
table(flight_data$carrier, flight_data$delay)
```

# Step 9: Redefine Delay Variable

```{r}
flight_data$delay_status <- ifelse(flight_data$delay == 1, "Delayed", "On Time")

table(flight_data$delay_status)
```

# Step 10: Summary of Major Variables

```{r}
summary(flight_data$distance)
summary(flight_data$schedtime)
summary(flight_data$deptime)
```

# Step 11: Histograms of Major Variables

```{r}
ggplot(flight_data, aes(x = distance)) +
  geom_histogram() +
  ggtitle("Histogram of Flight Distance")
```

# Step 12: Pie Chart (Flights Delayed)

```{r}
delay_count <- table(flight_data$delay_status)

pie(delay_count,
    main = "Delayed vs On-Time Flights")
```